Practical Hilbert space approximate Bayesian Gaussian processes for probabilistic programming

### 2022-03-07 Paper/Manuscript_Stats&Comp/manuscript_STCO.tex
minor edits

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions. However, the direct implementation entails a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a novel approach for low-rank approximate Bayesian Gaussian processes, based on a basis function approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance, and practical recommendations for how to select the number of basis functions and the boundary factor. Intuitive visualizations and recommendations, make it easier for users to improve approximation accuracy and computational performance.  We also propose diagnostics for checking that the number of basis functions and the boundary factor are adequate given the data. The approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\keywords{Gaussian process \and Low-rank Gaussian process \and Hilbert space methods \and Sparse Gaussian process \and Bayesian statistics \and Stan}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, domain of the prediction space, and properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the computational performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to choose valid and optimized values for these factors, improving computational performance without sacrificing modeling accuracy. We also propose  diagnostics to indicate whether the chosen values for the number of basis functions and the domain of the prediction space are adequate to model the data well.

@@

We start by illustrating how the number of basis functions $m$ and boundary factor $c$ influence the accuracy of the HSGP approximations individually. For this purpose, a set of noisy observations are drawn from an exact GP model with a squared exponential covariance function of length-scale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $c$ are fitted to this data. In this example, the length-scale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures~\ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictive mean and standard deviation of the estimated function as well as on the covariance function itself. For a sufficiently large fixed value of $c$, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation for both the posterior mean or uncertainty. It is seen that if the number of basis functions $m$ is too small, the estimated function tend to be overly smooth because the necessary high frequency components are missing. In general, the higher the degree of wigglyness of the function to be estimated, the larger number of  basis functions will be required. If $m$ is fixed to a sufficiently large value, Figure~\ref{fig2_Post_L} shows that $c$ affects the approximation of the mean mainly near the boundaries, while the approximation of the standard deviation is affected across the whole domain. What is more, the approximation error tends to be bigger for the standard deviation than for the mean.

@@

The component $f_4(t)$ represents the special days effects, modeled as a t-student prior model with 1 degree of freedom and variance $\tau^2$:
%
\begin{equation*}
f_4(t) \sim \text{t-student}(1,0,\tau^2).
\end{equation*}
%
%A horse-shoe prior allows for sparse distributed effects. Its global parameter $\tau$ pulls all the weights (effects) globally towards zero, while the thick half-Cauchy tails for the local scales $\lambda_t$ allow some of the weights to escape the shrinkage. Different levels of sparsity can be accommodated by changing the value of $\tau$: for large $\tau$ all the variables have very diffuse prior distributions with very little shrinkage towards zero, but letting $\tau \rightarrow 0$ will shrink all the weights $f_4(t)$ to zero \citep{piironen2016hyperprior}.

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,10)$ and $\mathrm{Normal}(0,2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variances $\bm{\alpha}=\{\alpha_1 \alpha_2, \alpha_3\}$, and length-scales $\bm{\ell}=\{\ell, \ell,\ell\}$, respectively. A $\mathrm{Normal}(0,0.1)$ prior distribution has been used for the standard deviation $\tau$ of the t-student distribution with 1 degree of freedom used to model $f_4$ (i.e., the special days effects).

@@

Modeling unknown functions using exact GPs is computationally intractable for a lot of applications. This problem becomes especially severe when performing full Bayesian inference using sampling-based methods. In this paper, a recent approach for a low-rank representation of stationary GPs, originally proposed by \citet{solin2018hilbert}, has been implemented and analyzed in detail. The method is based on a basis function approximation via Laplace eigenfunctions. The method has an attractive computational cost as it effectively approximates GPs by linear models, which is also an attractive property in modular probabilistic programming programming frameworks. The dominating cost per log density evaluation (during sampling) is $O(nm+m)$, which is a big benefit in comparison to $O(n^3)$ of a exact GP model. The obtained design matrix is independent of hyperparameters and therefore only needs to be constructed once, at cost $O(nm)$. All dependencies on the kernel and the hyperparameters are through the prior distribution of the regression weights. The parameters' posterior distribution is $m$-dimensional, where $m$ is usually much smaller than the number of observations $n$.

@@

The developed approximate GPs can be easily applied as modular components in probabilistic programming frameworks such as Stan in both Gaussian and non-Gaussian observation models. Using several simulated and real datasets, we have demonstrated the practical applicability and improved sampling efficiency, as compared to exact GPs, of the developed method. The main drawback of the approach is that its computational complexity scales exponentially with the number of input dimensions. Hence, choosing optimal values for the number of basis functions and the boundary factor, using the recommendations and diagnostics provided in Figures~\ref{fig6_7_relationships} and \ref{fig6_7_relationships}, is essential to avoid a excessive computational time especially in multivariate input spaces. However, in practice, input dimensionalities larger than three start to be quite computationally demanding even for moderately wiggly functions and few basis functions per input dimension. In these high dimensional cases, the proposed approximate GP methods may still be used for low-dimensional components in an additive modeling scheme but without modeling very high dimensional interactions, as complexity is linear with the number of additive components.

The obtained functional relationships between the key factors influencing the approximation not only help users to visually assess the accuracy of the method
but can also serve an automatic diagnostic tool, if appropriately implemented.
In this paper, we primarily studied the functional relationships for univariate inputs. Accordingly, investigating the functional relationships more thoroughly for multivariate inputs remains a topic for future research.

>>>>

\begin{abstract}
Gaussian processes are powerful non-parametric probabilistic models for stochastic functions. However, the direct implementation entails a complexity that is computationally intractable when the number of observations is large, especially when estimated with fully Bayesian methods such as Markov chain Monte Carlo. In this paper, we focus on a low-rank approximate Bayesian Gaussian processes, based on a basis function approximation via Laplace eigenfunctions for stationary covariance functions. The main contribution of this paper is a detailed analysis of the performance, and practical recommendations for how to select the number of basis functions and the boundary factor. Intuitive visualizations and recommendations, make it easier for users to improve approximation accuracy and computational performance.  We also propose diagnostics for checking that the number of basis functions and the boundary factor are adequate given the data. The approach is simple and exhibits an attractive computational complexity due to its linear structure, and it is easy to implement in probabilistic programming frameworks. Several illustrative examples of the performance and applicability of the method in the probabilistic programming language Stan are presented together with the underlying Stan model code.
\keywords{Gaussian process \and Low-rank Gaussian process \and Hilbert space methods \and Sparse Gaussian process \and Bayesian statistics \and Stan}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

While \citet{solin2018hilbert} have fully developed the mathematical theory behind this specific approximation of GPs, further work is needed for its practical implementation in probabilistic programming frameworks. In this paper, the interactions among the key factors of the method such as the number of basis functions, domain of the prediction space, and properties of the true functional relationship between covariates and response variable, are investigated and analyzed in detail in relation to the computational performance and accuracy of the method. Practical recommendations are given for the values of the key factors based on simple diagnostic values and intuitive graphical summaries that encode the recognized relationships. Our recommendations will help users to choose valid and optimized values for these factors, improving computational performance without sacrificing modeling accuracy. The proposed diagnostic can be used to check whether the chosen values for the number of basis functions and the domain of the prediction space are adequate to model the data well.

@@

We start by illustrating how the number of basis functions $m$ and boundary factor $c$ influence the accuracy of the HSGP approximations individually. For this purpose, a set of noisy observations are drawn from an exact GP model with a squared exponential covariance function of length-scale $\ell=0.3$ and marginal variance $\alpha=1$, using input values from the zero-mean input domain with half-range $S=1$. Several HSGP models with varying $m$ and $c$ are fitted to this data. In this example, the length-scale and marginal variance parameters used in the HSGPs are fixed to the true values of the data-generating model. 
Figures~\ref{fig1_Post_J} and \ref{fig2_Post_L} illustrate the individual effects of $m$ and $c$, respectively, on the posterior predictive mean and standard deviation of the estimated function as well as on the covariance function itself. For a sufficiently large fixed value of $c$, Figure \ref{fig1_Post_J} shows clearly how $m$ affects the accuracy on the approximation for both the posterior mean or uncertainty. It is seen that if the number of basis functions $m$ is too small, the estimated function tend to be overly smooth because the necessary high frequency components are missing. In general, the higher the degree of wigglyness of the function to be estimated, the larger number of  basis functions will be required. If $m$ is fixed to a sufficiently large value, Figure~\ref{fig2_Post_L} shows that $c$ affects the approximation of the mean mainly near the boundaries, while the approximation of the standard deviation is affected across the whole domain. The approximation error tends to be bigger for the standard deviation than for the mean.

@@

The component $f_4(t)$ represents the special days effects, modeled as a Student's $t$ prior model with 1 degree of freedom and variance $\tau^2$:
%
\begin{equation*}
f_4(t) \sim t(1,0,\tau^2).
\end{equation*}
%
%A horse-shoe prior allows for sparse distributed effects. Its global parameter $\tau$ pulls all the weights (effects) globally towards zero, while the thick half-Cauchy tails for the local scales $\lambda_t$ allow some of the weights to escape the shrinkage. Different levels of sparsity can be accommodated by changing the value of $\tau$: for large $\tau$ all the variables have very diffuse prior distributions with very little shrinkage towards zero, but letting $\tau \rightarrow 0$ will shrink all the weights $f_4(t)$ to zero \citep{piironen2016hyperprior}.

@@

The joint posterior parameter distributions are estimated by sampling using the dynamic HMC algorithm implemented in Stan \citep{carpenter2017stan,betancourt2017conceptual}. $\mathrm{Normal}(0,1)$, $\mathrm{Normal}(0,10)$ and $\mathrm{Normal}(0,2)$ prior distributions has been used for the observation noise $\sigma$, covariance function marginal variances $\bm{\alpha}=\{\alpha_1 \alpha_2, \alpha_3\}$, and length-scales $\bm{\ell}=\{\ell, \ell,\ell\}$, respectively. A $\mathrm{Normal}(0,0.1)$ prior distribution has been used for the standard deviation $\tau$ of the Student's $t$ distribution with 1 degree of freedom used to model $f_4$ (i.e., the special days effects).

@@

Modeling unknown functions using exact GPs is computationally intractable in many applications. This problem becomes especially severe when performing full Bayesian inference using sampling-based methods. In this paper, a recent approach for a low-rank representation of stationary GPs, originally proposed by \citet{solin2018hilbert}, has been analyzed in detail. The method is based on a basis function approximation via Laplace eigenfunctions. The method has an attractive computational cost as it effectively approximates GPs by linear models, which is also an attractive property in modular probabilistic programming programming frameworks. The dominating cost per log density evaluation (during sampling) is $O(nm+m)$, which is a big benefit in comparison to $O(n^3)$ of a exact GP model. The obtained design matrix is independent of hyperparameters and therefore only needs to be constructed once, at cost $O(nm)$. All dependencies on the kernel and the hyperparameters are through the prior distribution of the regression weights. The parameters' posterior distribution is $m$-dimensional, where $m$ is usually much smaller than the number of observations $n$.

@@

The developed approximate GPs can be easily applied as modular components in probabilistic programming frameworks such as Stan in both Gaussian and non-Gaussian observation models. Using several simulated and real datasets, we have demonstrated the practical applicability and improved sampling efficiency, as compared to exact GPs, of the developed method. The main drawback of the approach is that its computational complexity scales exponentially with the number of input dimensions. Hence, choosing optimal values for the number of basis functions and the boundary factor, using the recommendations and diagnostics provided in Figures~\ref{fig6_7_relationships} and \ref{fig6_7_relationships}, is essential to avoid a excessive computational time especially in multivariate input spaces. However, in practice, input dimensionalities larger than three start to be computationally demanding even for moderately wiggly functions and few basis functions per input dimension. In these high dimensional cases, the proposed approximate GP methods may still be used for low-dimensional components in an additive modeling scheme but without modeling very high dimensional interactions, as complexity is linear with the number of additive components.

The obtained functional relationships between the key factors influencing the approximation not only help users to visually assess the accuracy of the method
but can also serve an automatic diagnostic tool.
In this paper, we primarily studied the functional relationships for univariate inputs. Accordingly, investigating the functional relationships more thoroughly for multivariate inputs remains a topic for future research.
